---
title: "ISLR Chap 4 Notes"
author: "M. E. Waggoner"
date: "February 24, 2019"
output: 
  pdf_document:
    toc: yes
    toc_depth: 5
    pandoc_args: [
      "--number-sections",
      "--number-offset=4"
    ]
---

```{r setup, include=FALSE}
require(tidyverse)    # various
require(texreg)       # screenreg()
require(ggfortify)    # autoplot()
require(knitr)        # kable()
require(broom)        # tidy()
require(MASS)         # lda(), qda()
require(car)          # knn()

options(digits = 4)
```


$\setcounter{section}{3}$

# Classification

## An Overview of Classification

Note:  The section numbers of this document match up with ISLR.  The equation numbers match up with the section, but you need to append "4." to the number.

A factor variable can be called *qualitative*, *categorical*, or other names that indicate that it takes on a finite number of *classes*, *levels*, or *factors*.  When the output variable is qualitative, the process of prediction is called *classification* rather than regression.

This section of ISLR covers four methods of classification:  

* logistic regression,

* linear discriminant analysis (LDA),

* quadratic discriminant analysis (QDA), and

* K-nearest neighbors (KNN).

Other classifiers are covered in Chapters 7, 8, and 9.

#### LDA illustration

```{r}
midge.train <- as.tibble(read.csv("C:/Users/murphy.waggoner/Documents/GitHub/Math-308-Adv-Data-Modeling/ISLR Chapter 01/ISLR Chapter 1 Midge.txt", 
                                  header = TRUE))

str(midge.train)

meanWL <- mean(midge.train$WingLength)
meanAL <- mean(midge.train$AntennaLength)

mref <- 9.66/14.78
bref <- meanAL - mref*meanWL

midge.train %>% 
  ggplot(aes(x = WingLength, y = AntennaLength, 
             col = Type, shape = Type)) + 
  geom_point() +
  geom_point(aes(x = meanWL, y = meanAL), colour="black", size = 2) +
  geom_abline(slope = mref, intercept = bref) -> baseplot

midge.train %>% 
  mutate(bperp = AntennaLength - (-1/mref)*WingLength) -> midge.train

for(i in 1:nrow(midge.train)){
  baseplot + geom_abline(slope = -1/mref, intercept = midge.train$bperp[[i]] ,
                         alpha = 0.2) -> baseplot
}

midge.train %>% 
  mutate(xproj = (bref - bperp)/(-1/mref - mref),
         yproj = mref*(xproj) + bref) -> midge.train


baseplot + geom_point(aes(x = xproj, y = yproj, color = Type), size = 2, data = midge.train)

midge_lda <- lda(Type ~ WingLength + AntennaLength, data = midge.train)

midge_lda

plot(midge_lda)
```



